---
title: "P8105 Homework 6"
author: "Maria Serafini"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(modelr)
library(mgcv)
library(p8105.datasets)

knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
  )

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
  )

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```


## Problem 1

The code below imports the `homicide-data` dataset from the Washington Post. This dataset consists of individual homicide records collected from 50 large U.S. cities. Each row represents a single homicide case, and the variables provide information about the victim, the circumstances of the homicide, and the status of the investigation. Key variables include the `city` and `state` in which the homicide occurred, the `reported_date` of the homicide, the approximate geographic location of the homicide given by `longitude` and `latitude`, and `disposition` which represents the investigative status of the case. 

```{r}
homicide_df <- read_csv("data/homicide-data.csv")
```


The code below creates a `clean_homicide_df` dataset, which includes a city-level summary of the homicide data. The `city` and `state` variables are combined into a single`city_state` variable. A binary `solved` variable is created to indicate whether the homeicide was resolved by arrest. Cities that do not report the race of the homicide victim are omitted. A misclassified city  “Tulsa, AL” is removed.Victim race is limited to "White" or "Black" for analysis purposes. 

```{r}
clean_homicide_df <- 
  homicide_df |> 
  unite(city_state, city, state, sep = ", ") |> 
  mutate(
    solved = as.numeric(disposition == "Closed by arrest"), 
    victim_age = as.numeric(victim_age)
  ) |> 
  filter(!city_state %in% c("Dallas, TX", "Phoenix, AZ", "Kansas City, MO", 
                         "Tulsa, AL")) |> 
  filter(
    victim_race %in% c("White", "Black")
  ) 
```


The code below fits a logistic regression to the Baltimore, MD data with solved vs unsolved as the outcome and victim age, sex and race as predictors. 

```{r}
baltimore_df <- 
  clean_homicide_df |> 
  filter(
    city_state == "Baltimore, MD"
  ) |> 
  select(solved, victim_age, victim_sex, victim_race)

baltimore_logfit = 
  baltimore_df |> 
  glm(solved ~ victim_age + victim_race + victim_sex, data = _, family = binomial()) 

baltimore_results <- 
  baltimore_logfit |> 
  broom::tidy(conf.int = TRUE, exponentiate = TRUE) |> 
  filter(term == "victim_sexMale") |> 
  select(term, 
         OR = estimate,
         CI_lower = conf.low,
         CI_upper = conf.high, 
         p.value)

baltimore_results |> 
  knitr::kable(digits = 3)
```


In Baltimore, MD, the adjusted odds ratio for solving homicides comparing male to female victims is approximately 0.43 (95% CI: 0.32, 0.56), indicating that homicides with male victims have substantially lower odds of being solved than those with female victims, after adjusting for age and race.

```{r}
city_glm_results <- 
  clean_homicide_df |> 
  nest(data = -city_state) |> 
  mutate(
    fits = map(data, \(df) glm(solved ~ victim_age + victim_race + victim_sex, data = df, 
                               family = binomial())), 
    results = map(fits, broom::tidy, conf.int = TRUE, exponentiate = TRUE)
  ) |> 
  unnest(results) |> 
  filter(term == "victim_sexMale") |> 
  select(
    city_state, 
    OR = estimate,
    CI_lower = conf.low,
    CI_upper = conf.high,
    p.value
  )

city_glm_results |> 
  knitr::kable(digits = 3)
```


Overall, the plot below shows that in most cities, homicides involving male victims have lower adjusted odds of being solved than those involving female victims, and in roughly half of the cities this difference is statistically significant. A small number of cities have odds ratios above 1, but wide confidence intervals that include 1, indicating substantial uncertainty rather than strong evidence of a reversal of this pattern.

```{r, fig.width=10, fig.height=10}
city_glm_results |> 
  mutate(city_state = fct_reorder(city_state, OR)) |> 
  ggplot(aes(x = OR, y = city_state)) + 
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_errorbarh(aes(xmin = CI_lower, xmax = CI_upper), height = 0.2) +
  geom_point() + 
  labs(
    title = "Adjusted Odds of Resolved Homicide in Male vs Female Victims By City",
    x = "Adjusted Odds Ratio (Male vs Female Victims)",
    y = "City"
  ) +
  theme(axis.text.y = element_text(size = 8)) 
```



## Problem 2

The code below imports the Central Park weather data from p8105.datasets. This dataset contains daily meteorological observations recorded in New York City, including minimum and maximum temperature (`tmin`, `tmax`) and total daily precipitation (`prcp`). 

```{r}
data("weather_df")
```


The code below generates 5000 bootstrap samples from the original dataset and fits the linear regression model $tmax = \beta_0 + \beta_1\, tmin + \beta_2\, prcp + \varepsilon$ to extract the estimated quantities $\widehat{r}^2$ and $\widehat{\beta}_1 / \widehat{\beta}_2$. For each bootstrap sample, the model is re-fit, and the two quantities of interest are extracted using `broom::glance()` for $\widehat{r}^2$ and `broom::tidy()` for $\widehat{\beta}_1 / \widehat{\beta}_2$.

```{r}
set.seed(1)

boot_weather <- 
  weather_df |> 
  modelr::bootstrap(n = 5000) |> 
  mutate(
    models = map(strap, \(df) lm(tmax ~ tmin + prcp, data = df)), 
    results = map(models, broom::tidy), 
    glance = map(models, broom::glance)
  ) 


boot_rsquared <- 
  boot_weather |> 
  select(.id, glance) |> 
  unnest(glance) |> 
  transmute(.id, r2 = r.squared)

boot_ratio <- 
  boot_weather |> 
  select(.id, results) |> 
  unnest(results) |> 
  select(.id, term, estimate) |> 
  filter(term %in% c("tmin", "prcp")) |> 
  pivot_wider(
    names_from = term,
    values_from = estimate
  ) |> 
  mutate(
    beta_ratio = tmin / prcp) |> 
  select(.id, beta_ratio)
 
```


The plots below show the bootstrap sampling distributions for both estimated quantities. The distribution of $\widehat{r}^2$ is tightly concentrated, reflecting the strong linear relationship between minimum and maximum temperature. In contrast, the bootstrap distribution of the coefficient ratio $\widehat{\beta}_1 / \widehat{\beta}_2$ is much more variable and noticeably left-skewed. This increased variability and skewness are consistent with the relatively small and unstable estimated effect of precipitation in the regression model.

```{r}
boot_summary <- 
  boot_rsquared |> 
  inner_join(boot_ratio, by = ".id")

boot_summary |> 
  pivot_longer(
    cols = c(r2, beta_ratio),
    names_to = "quantity", 
    values_to = "value", 
  ) |> 
  mutate(
    quantity = recode(
      quantity, 
      "r2" = "hat(r)^2",
      "beta_ratio" = "hat(beta)[1] / hat(beta)[2]")
  ) |> 
  ggplot(aes(x = value)) +
  geom_density(color = "blue") + facet_wrap(~ quantity, scales = "free", labeller = label_parsed) +
  labs(
    title = expression(
      "Bootstrap distributions of " * hat(r)^2 * " and " * hat(beta)[1] / hat(beta)[2]
    ),
    x = "Bootstrap Estimate", 
    y = "Density"
  )
```

The code below identifies the 2.5% and 97.5% quantiles of each bootstrap distribution to construct 95% confidence intervals for the estimated quantities.

```{r}
boot_summary |> 
  summarize(
    r2_lower = quantile(r2, 0.025), 
    r2_upper = quantile(r2, 0.975), 
    ratio_lower = quantile(beta_ratio, 0.025), 
    ratio_upper = quantile(beta_ratio, 0.975)
  )
```


## Problem 3

The code below imports the birthweight dataset, which contains detailed information on approximately 4,000 infants and their mothers. Each row represents a single birth, and the dataset includes variables describing infant characteristics (e.g., sex, birthweight, head circumference, and length), maternal and paternal demographic factors, maternal health indicators, and pregnancy-related behaviors. Key variables include `bwt` (birthweight in grams), `blength` and `bhead` (infant length and head circumference at birth), `gaweeks` (gestational age in weeks), and several maternal characteristics such as pre-pregnancy BMI, weight gain during pregnancy, smoking, and age at delivery.

```{r}
birthweight_df <- read_csv("data/birthweight.csv")
```


To prepare the dataset for regression analysis, descriptive variable names were applied and several numeric variables were converted to factors to better reflect their categorical structure. The dataset was then assessed for completeness by computing the number of missing values in each variable. As shown below, all variables contain zero missing values so no additional filtering is required.

```{r}
birthweight_df <-
  birthweight_df |> 
  rename(
    sex = babysex, 
    head_circumference = bhead, 
    birth_length = blength,
    birth_wt = bwt, 
    mom_wt_delivery = delwt, 
    monthly_income = fincome, 
    dad_race = frace, 
    mom_race = mrace, 
    gestational_age = gaweeks, 
    malformations = malform,
    mom_age_menarche = menarche, 
    mom_height = mheight, 
    mom_age_delivery = momage, 
    count_past_lbwt = pnumlbw,
    count_past_small_ga = pnumsga, 
    mom_pre_bmi = ppbmi, 
    mom_pre_wt = ppwt, 
    count_smoke = smoken, 
    mom_wt_gain = wtgain
  ) |> 
  mutate(
    sex = factor(sex, levels = c(1,2), labels = c("Male", "Female")), 
    dad_race = factor(
      dad_race,
      levels = c(1,2,3,4,8,9),
      labels = c("White", "Black", "Asian", "Puerto Rican", "Other", "Unknown")), 
    mom_race = factor(
      mom_race, 
      levels = c(1,2,3,4,8),
      labels = c("White", "Black", "Asian", "Puerto Rican", "Other")), 
    malformations = factor(
      malformations, 
      levels = c(0, 1),
      labels = c("absent", "present"))
    ) 


birthweight_df |> 
  summarise(across(everything(), ~ sum(is.na(.))))
```


The code below fits a proposed linear regression model for birthweight that incorporates a set of biologically and clinically relevant predictors. This model includes infant head circumference, birth length, gestational age, maternal pre-pregnancy BMI, maternal weight gain during pregnancy, number of cigarettes smoked per day, infant sex, paternal race, and parity. These variables were selected to capture fetal size at birth, duration of gestation, maternal health and pregnancy-related behaviors, and demographic factors that may influence birth outcomes.

```{r, proposed_model}
birth_wt_model <- 
  lm(birth_wt ~ head_circumference + birth_length + gestational_age + mom_pre_bmi +
                         mom_wt_gain + count_smoke + sex + dad_race + parity, data = birthweight_df)

birth_wt_model |> 
  broom::tidy()

birth_wt_model |> 
  broom::glance()
```


The code below creates a residuals-versus-fitted plot using `add_predictions()` and `add_residuals()` to evaluate model assumptions. The plot shows residuals centered around zero with no strong evidence of curvature or heteroskedasticity, suggesting that the linear model form is broadly appropriate for these predictors.

```{r}
birthweight_df |> 
  add_predictions(birth_wt_model) |> 
  add_residuals(birth_wt_model) |> 
  ggplot(aes(x = pred, y = resid)) +
  geom_point(alpha = 0.4) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x = "Fitted Birthweight (Grams)",
    y = "Residuals",
    title = "Residuals vs Fitted Values for Proposed Model"
  )
```


For comparison, the code below fits a main-effects model including only birth length and gestational age as predictors. 

```{r main_effects_model}
main_effects_model <- 
  lm(birth_wt ~ birth_length + gestational_age, data = birthweight_df) 
  
main_effects_model |> 
  broom::tidy()
```


For comparison, the code below fits an interaction model including head circumference, birth length, sex, and all corresponding interaction terms (including the three-way interaction).

```{r interactions_model}
interactions_model <- 
  lm(birth_wt ~  birth_length * head_circumference * sex, data = birthweight_df) 

interactions_model |> 
  broom::tidy()
```


To compare the predictive performance of the three models, the code below applies `crossv_mc` with 100 random training-test splits. For each split, all three models were fit to the training data, and the root mean squared error (RMSE) was computed on the corresponding test set. 

```{r}
cross_validation <-  
  crossv_mc(birthweight_df, 100) |> 
  mutate(
    train = map(train, as_tibble),
    test = map(test, as_tibble)) |> 
  mutate(
    proposed_model     = map(train, \(df)  
                             lm(birth_wt ~ head_circumference + birth_length + 
                                gestational_age + mom_pre_bmi + mom_wt_gain + count_smoke 
                                + sex + dad_race + parity, data = df)),
    main_effects_model = map(train, \(df) 
                             lm(birth_wt ~ birth_length + gestational_age, data = df)),
    interactions_model = map(train, \(df) 
                             lm(birth_wt ~  birth_length * head_circumference * sex, data = df)) 
    ) |> 
  mutate(
    rmse_proposed     = map2_dbl(proposed_model, test, \(mod, df) rmse(model = mod, data = df)),
    rmse_main_effects = map2_dbl(main_effects_model, test, \(mod, df) rmse(model = mod, data = df)),
    rmse_interactions = map2_dbl(interactions_model, test, \(mod, df) rmse(model = mod, data = df))
    )
```


The violin plot below summarizes the distribution of cross-validated RMSE for each model across the 100 splits. This provides a direct comparison of out-of-sample predictive accuracy and allows us to assess whether the additional predictors or interaction terms meaningfully improve performance relative to the simpler models.

```{r}
cross_validation |> 
  select(starts_with("rmse")) |> 
  pivot_longer(
    everything(),
    names_to = "model", 
    values_to = "rmse",
    names_prefix = "rmse_") |> 
  mutate(model = fct_inorder(model)) |> 
  ggplot(aes(x = model, y = rmse)) + geom_violin()
```


Across the 100 cross-validation splits, the proposed model shows the lowest and most stable prediction error, with RMSE values generally centered around the mid-270s. The interaction model performs slightly worse, with RMSEs clustered around the low- to mid-280s, indicating that adding the three-way and two-way interactions between infant sex, head circumference, and birth length does not meaningfully improve predictive accuracy. The main-effects model using only birth length and gestational age performs substantially worse than the other two models, with RMSE values in the 320–350 range. Overall, the proposed model provides the strongest out-of-sample predictive performance among the three, suggesting that incorporating additional maternal and demographic predictors improves prediction of birthweight beyond infant size and gestational age alone, while the complex interaction structure does not offer additional benefit.